---
title: "Reproducible Research: NOAA Storm Database analysis"
output: 
  html_document:
    keep_md: true
---

## Synopsis
In this analysis, we are trying to answer the following questions:
1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?


So we conclude that:
According to population health, tornados resulted in the most impact on event fatality, injury (by total number of cases). As for economic consequences, flooding appeared to have the most impact.

## Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

The basic goal of this research is to explore the NOAA Storm Database and answer some basic questions about severe weather events.

Our data analysis address to the following questions:

1) Across the United States, which types of events are most harmful with respect to population health?
2) Across the United States, which types of events have the greatest economic consequences?


## Data Processing

The data for this research come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from web site:


[Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) ~47Mb, 902297 obs., 37 vars.


[Data description](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)


The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.


So, we just load the data and extract an appropriate vars:


- EVTYPE - types of event (e.g. tornado)
- FATALITIES - fatalities count
- INJURIES - injuries count
- PROPDMG - estimation of property damage, $


```{r, cache=TRUE}

data<-read.csv("repdata_data_StormData.csv.bz2")
data<-data[,c(8,23,24,25)] #extract EVTYPE, FATALITIES, INJURIES, PROPDMG
data$EVTYPE<-toupper(data$EVTYPE) #uppercase correction
summary(data)

```


## Results


According to the first question, let's see how total fatalities and injuries distributed by event type. 


Top-10 are:
```{r}
f<-aggregate(FATALITIES ~ EVTYPE, data, sum)
i<-aggregate(INJURIES ~ EVTYPE, data, sum)

f<-f[f$FATALITIES > 0,]   #zero-values ex.
i<-i[i$INJURIES > 0,]

f<-f[order(-f$FATALITIES),] #sort
i<-i[order(-i$INJURIES),]

head(f, n = 10)
head(i, n = 10)

```

Let's see this on plot:

```{r}
f10<-head(f, n = 10)
i10<-head(i, n = 10)

f10<-f10[order(f10$FATALITIES),] #sort
i10<-i10[order(i10$INJURIES),]

par(mfcol=c(1,2), mar=c(2,7,1,1))

barplot(height=f10$FATALITIES, names=tolower(f10$EVTYPE), col=factor(f10$EVTYPE), horiz=T, las=1, main = "Total fatalities")

barplot(height=i10$INJURIES, names=tolower(i10$EVTYPE), col=factor(i10$EVTYPE), horiz=T, las=1, main = "Total injuries")

```

According to the second question, let's see how total damage distributed by event type. 
Top-10 are:
```{r}
p<-aggregate(PROPDMG ~ EVTYPE, data, sum)

p<-p[p$PROPDMG > 0,]   #zero-values ex.

p<-p[order(-p$PROPDMG),] #sort

head(p, n = 10)

```

Let's see this on plot:

```{r}
p10<-head(p, n = 10)

p10<-p10[order(p10$PROPDMG),] #sort

par(mar=c(2,7,1,1))

barplot(height=p10$PROPDMG, names=tolower(p10$EVTYPE), col=factor(p10$EVTYPE), horiz=T, las=1, main = "Total property damage, $")

```

Now, let's see on top-5 average values:

```{r, cahe=TRUE}
f<-aggregate(FATALITIES ~ EVTYPE, data, mean)
i<-aggregate(INJURIES ~ EVTYPE, data, mean)
p<-aggregate(PROPDMG ~ EVTYPE, data, mean)

p<-p[order(-p$PROPDMG),] #sort
f<-f[order(-f$FATALITIES),] 
i<-i[order(-i$INJURIES),]

p5<-head(p, n = 5)
f5<-head(f, n = 5)
i5<-head(i, n = 5)

f5
i5
p5

p5<-p5[order(p5$PROPDMG),] #sort
f5<-f5[order(f5$FATALITIES),] 
i5<-i5[order(i5$INJURIES),]

par(mfcol=c(3,1), mar=c(2,7,1,1))

barplot(height=f5$FATALITIES, names=tolower(f5$EVTYPE), col=factor(f5$EVTYPE), horiz=T, las=1, main = "Average fatalities")

barplot(height=i5$INJURIES, names=tolower(i5$EVTYPE), col=factor(i5$EVTYPE), horiz=T, las=1, main = "Average injuries")

barplot(height=p5$PROPDMG, names=tolower(p5$EVTYPE), col=factor(p5$EVTYPE), horiz=T, las=1, main = "Average property damage, $")

```

## Discussion


On average tornado, cold and snow, and wild fires are the most influence types with respect to population health. But the most expensive types are some kinds of flood and coastal erosion.
This context may be very different by States.


You may access that puplication on my [RPubs](https://rpubs.com/Ellariel/noaa)


